---
title: "NYPD Shooting Incident data analysis"
author: "Yujung Teng"
date: "2021/7/29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Hi There. I'm going to show you my analysis process and results of NYPD Shooting Incident.

First at all. English is not my Mother tongue. I want to declare that I have no malice against any race, gender, or country. If there is any wrong expression in English, please forgive me.


## 1. read library and data

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(zoo)
library(rpart)
```

Download data by url will be easy to Reproduce, from https://catalog.data.gov/dataset and searching for
‘NYPD Shooting Incident Data (Historic)’.

```{r message=FALSE, warning=FALSE}
url_in = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD_data = read_csv(url_in)
```

## 2. Tidy & have a look with data

Do the summary and We can see the OCCUR_DATE & OCCUR_TIME are not in the right type

So let's Change them to date & time types respectively
```{r}
NYPD_data = NYPD_data%>%
    mutate(OCCUR_DATE=mdy(OCCUR_DATE))

NYPD_data = NYPD_data%>%
    mutate(OCCUR_TIME=hms(OCCUR_TIME))
```


Now there are in correct type. And we can get the data date range is 2006-01-01 to 2020-12-31
```{r}
summary(NYPD_data)
```


Let's see how many incident occurred in each borough.
```{r}
NYPD_data %>% group_by(BORO) %>% summarise(COUNT=n()) %>% arrange(desc(COUNT))
```
As we all know, different borough have different population.
Some borough has more incidents just because it has larger population.
To avoid this bias, I'm going to download the population data of New-York

## 3. Find additional data. Transform and Merge

That I found is the population data of New-York from NYC open data, which is a Official website. We can believe the data from it will be correct.

Again, download data from url for easily reproduce.

```{r}
url_in ="https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv?accessType=DOWNLOAD"
NY_POPULATION = read_csv(url_in)
```

We can see each borough's population in decade in summary.
```{r}
summary(NY_POPULATION)
```

The Borough columns have all the Borough in New-York. And have we can also get the NYC Total.
```{r}
NY_POPULATION$Borough 
```



Now I am going to merge these two data set.

Since Our incident data which date range is 2006-01-01 to 2020-12-31. We only need population in 2000~2020.

I transformed the population data which columns = (Borough , year of decade, population).
And I transformed a new data frame for NYC Total.
```{r}
NY_POPULATION_DECADE = NY_POPULATION%>% 
  select('Borough','2000','2010','2020')%>% 
  mutate(Borough = toupper(Borough))%>% 
  gather(YEAR_DECADE,POPULATION, `2000`:`2020`)%>% 
  rename("BORO"="Borough")%>% 
  mutate(YEAR_DECADE = as.integer(YEAR_DECADE))

NY_POPULATION_TOTAL = NY_POPULATION_DECADE %>% 
  filter(BORO=="NYC TOTAL") %>% 
  rename("ALL_NY_POPULATION"="POPULATION")%>%
  select(-(BORO))
```


For the incidence data,What I am interesting is when the incident occur.

So I Select the date & time columns, and create some new columns like year,month,day,hour.
```{r}
NYPD_TIME_BORO = NYPD_data %>% select(OCCUR_DATE,OCCUR_TIME,BORO,STATISTICAL_MURDER_FLAG)
NYPD_TIME_BORO = NYPD_TIME_BORO%>%
    mutate(OCCUR_YEAR = year(OCCUR_DATE),
          #YEAR_DECADE = round_any(OCCUR_YEAR, 10, f = ceiling),
           YEAR_DECADE = as.integer(paste(substr(as.character(OCCUR_YEAR),0,3),'0',sep='')),
           OCCUR_MONTH = month(OCCUR_DATE),
           OCCUR_DAY = day(OCCUR_DATE),
           OCCUR_WDAY = wday(OCCUR_DATE),
           OCCUR_YEARMON =as.yearmon(OCCUR_DATE),
           OCCUR_HOUR = hour(OCCUR_TIME)   
)
```


Merge incident and population data.
```{r}
NYPD_TIME_BORO = left_join(NYPD_TIME_BORO, NY_POPULATION_DECADE
                           , by = c("BORO" = "BORO", "YEAR_DECADE" = "YEAR_DECADE"))

NYPD_TIME_BORO = left_join(NYPD_TIME_BORO, NY_POPULATION_TOTAL
                           , by = c("YEAR_DECADE" = "YEAR_DECADE"))
```


Make sure there is no NA in the data.
```{r}
colSums(is.na(NYPD_TIME_BORO))
```
Now we can start to analysis.


## 4. Analysation & Visualization

First,let's look the Shooting crime rate for each year

Shooting crime rate = 1000 * Year_COUNT/POPULATION

We can see the Shooting crime rate is getting down but have a great increase in 2020
```{r,fig.width=6, fig.height=2}
NYPD_YEAR_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_YEAR) %>% 
  summarise(COUNT=n(),ALL_NY_POPULATION=max(ALL_NY_POPULATION)) %>% 
  mutate(rate_shoot = 1000 * COUNT/ALL_NY_POPULATION) %>% 
  arrange(OCCUR_YEAR)

ggplot(NYPD_YEAR_COUNT
       ,aes(x=OCCUR_YEAR,y=rate_shoot))+geom_point()+geom_line()
```

let's see with smaller period, group by YEARMONTH.

Interesting, it has periodic fluctuations, like monthly change
```{r,fig.width=6, fig.height=2}
NYPD_YM_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_YEARMON) %>% 
  summarise(COUNT=n(),ALL_NY_POPULATION=max(POPULATION)) %>% 
  mutate(rate_shoot = 1000 * COUNT/ALL_NY_POPULATION) %>% 
  arrange(OCCUR_YEARMON)

ggplot(NYPD_YM_COUNT
       ,aes(x=OCCUR_YEARMON,y=rate_shoot))+geom_point()+geom_line()
```

let's see all the data group by month.

Obviously the peak around July and August,And the valley around February

```{r,fig.width=6, fig.height=2}
NYPD_YEAR_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_MONTH) %>% 
  summarise(COUNT=n(),ALL_NY_POPULATION=max(ALL_NY_POPULATION)) %>% 
  mutate(rate_shoot = 1000 * COUNT/ALL_NY_POPULATION) %>% 
  arrange(OCCUR_MONTH)

ggplot(NYPD_YEAR_COUNT
       ,aes(x=OCCUR_MONTH,y=rate_shoot))+geom_point()+geom_line()+ scale_x_continuous(breaks = seq(1, 12, by = 1))
```


let's see the each borough if they have the same periodic fluctuations

The answer is yes.

```{r warning=FALSE}
NYPD_MONTH_BORO_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_MONTH,BORO) %>% 
  summarise(COUNT=n(),POPULATION=max(POPULATION), .groups = 'drop') %>% 
  mutate(rate_shoot = 1000 * COUNT/POPULATION) %>% 
  arrange(OCCUR_MONTH)


ggplot(NYPD_MONTH_BORO_COUNT
       ,aes(x=OCCUR_MONTH,y=rate_shoot)
       )+geom_point()+geom_line()+facet_wrap(~BORO)+ scale_x_continuous(breaks = seq(1, 12, by = 1))
```

And I also print the incidence data group by hours and weekday.

By hours,We can see most of incidence occurred at afternoon and night.
```{r}
NYPD_HOUR_BORO_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_HOUR,BORO) %>% 
  summarise(COUNT=n(),POPULATION=max(POPULATION), .groups = 'drop') %>% 
  mutate(rate_shoot = 1000 * COUNT/POPULATION) %>% 
  arrange(OCCUR_HOUR)


ggplot(NYPD_HOUR_BORO_COUNT
       ,aes(x=OCCUR_HOUR,y=rate_shoot)
       )+geom_point()+geom_line()+facet_wrap(~BORO)+ scale_x_continuous(breaks = seq(0, 23, by = 2))

```

By Weekday We can see most of incidence occurred in Saturday & Sunday (1 = Sunday ,7 = Saturday)
```{r}
NYPD_WDAY_BORO_COUNT = NYPD_TIME_BORO %>% 
  group_by(OCCUR_WDAY,BORO) %>% 
  summarise(COUNT=n(),POPULATION=max(POPULATION), .groups = 'drop') %>% 
  mutate(rate_shoot = 1000 * COUNT/POPULATION) %>% 
  arrange(OCCUR_WDAY)


ggplot(NYPD_WDAY_BORO_COUNT
       ,aes(x=OCCUR_WDAY,y=rate_shoot)
       )+geom_point()+geom_line()+facet_wrap(~BORO)+ scale_x_continuous(breaks = seq(1, 7, by = 1))
```

## 5. Modeling

Let's try to make a model for the data group by hour and each borough 
```{r}
NYPD_HOUR_BORO_rate = NYPD_TIME_BORO %>% 
  group_by(OCCUR_HOUR,BORO) %>% 
  summarise(COUNT=n(),
            POPULATION=max(POPULATION),
            death= sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>% 
  mutate(rate_shoot = 1000 * COUNT/POPULATION,
         rate_death = death/COUNT,
         ) %>% 
  arrange(OCCUR_HOUR)

```

I used CART for modeling, and print the Tree
```{r,fig.width=7, fig.height=7}
mod = rpart(rate_shoot ~ OCCUR_HOUR , data=NYPD_HOUR_BORO_rate)
plot(mod)
text(mod , cex =.7)
```

let's do the prediction.

We can see the Result is match
```{r}
NYPD_HOUR_BORO_rate = NYPD_HOUR_BORO_rate%>%
  mutate(pred= predict(mod,NYPD_HOUR_BORO_rate %>% select(OCCUR_HOUR)))


NYPD_HOUR_BORO_rate%>% 
  ggplot()+
  geom_point(aes(x = OCCUR_HOUR, y= rate_shoot),color = "blue")+
  geom_line(aes(x = OCCUR_HOUR, y= pred),color = "red")
```


## Conclusion
There must be a clear relationship between the date and time of the shooting. This is an area where further investigation into the circumstances and factors that led to these incidents may be required.

## Bias
1. Maybe shooting incidents occur at dawn is rarely reported
2. Maybe people will be avaliable to call the police at the weekends but not in working day.

Thank you ! 